### 2023-06-23 python_excel_ite.tex
working on intro

Despite numerous calls for and predictions of the eradication of spreadsheets from the business world, Excel and other spreadsheets are alive and well. Yes, spreadsheets have well known limitations. Spreadsheets are not databases and probably should not be used as one, especially for relational data models. There are better data visualization and exploration tools. There are better dashboarding tools that are more easily maintained in a production environment. Spreadsheet errors are a huge problem. Version control and documentation present challenges. There are data size limitations and external data links present their own set of challenges. But, spreadsheets are still very well suited to ``what if?'' type modeling due to their flexibility, numerous built in functions, and their integrated sensitivity analysis, goal seeking and visualization capabilities. Excel is widely used for building and using models of business problems
to explore the impact of various model inputs on key outputs. Built in
tools such as \href{https://support.microsoft.com/en-us/office/calculate-multiple-results-by-using-a-data-table-e95e2487-6ca6-4413-ad12-77542a5ea50b}{Data
	Tables} and
\href{https://support.microsoft.com/en-us/office/use-goal-seek-to-find-the-result-you-want-by-adjusting-an-input-value-320cb99e-f4a4-417f-b1c3-4f369d6e66c7}{Goal
	Seek} are well known to power spreadsheet modelers.

But, the analytics world has been changing. The advent of programmatic analytics using tools like R and Python have changed the practice of analytics as well as the teaching landscape \cite{bibid}. Limitations of spreadsheets with respect to reproducibility and automation have certainly played a role.  Additionally, both R and Python are free, open source, and have a rich set of analytics related libraries that have contributed to their growing popularity. Python, being a general purpose programming language, also offers advantages when moving models and analytical analysis pipelines from the prototype stage into production. Enormous data science ecosystems have emerged around both R and Python providing a wealth of learning resources for both new and experienced analytics students and professionals.

Numerous R and Python based data science courses have been developed and offered at universities and online learning platforms such as Coursera, EdX, Udemy and others. Some of these courses have made a large amount of their course content freely available for self-learners and for other educators. At our institution, we created a course called Practical Computing for Data Analytics (PCDA) in 2015 that introduced both R and Python (along with a little Linux) for use in data analytics work. The course is taught using a combination of R Markdown documents \cite{bibid} and Jupyter notebooks \cite{bibid}. Jupyter notebooksLiterate computing \cite{Jupyter_proposal}. Educational narratives that support iterative exploration and development. The ability to weave standard language with formal programming languages is critical to creating the desired educational artifact. Jupyter used by many universities as a key component of their educational technology stack.All of the course content is freely available from a publicly accessible course web site \url{URL}.

The PCDA course focuses mostly on data wrangling, exploratory data analysis (EDA), and statistical/machine learning based predictive models. It does include a Python based simulation of the Monte Hall 3-door problem. In teaching EDA using tools like ggplot2, students are challenged to think about how they would go about trying to use Excel to create some of the plots that are easy to do with R and Python based plotting libraries. There is also some limited coverage of using Python to open Excel files and do some sheet manipulation. However, there was no real attempt to do the kind of modeling that was the focus of spreadsheet based modeling textbooks.

Python has continued to make inroads to analytical territory that was formerly dominated by Excel. The widely used Python data analysis library, pandas, was created by Wes McKinney while working as a quantitative financial analyst \cite{bibid}.

Python is widely used for data wrangling, visualization and predictive modeling using various flavors of regression and tree based models. The blog \href{https://pbpython.com/}{Practical Business Python} written primarily by Chris Moffitt contributed greatly to making Python more accessible to business analytics practitioners through high quality posts that were extremely practical and useful. But, it got me to thinking. What about those things for which Excel is well suited such as building formula based models and doing sensitivity analysis on these models? What would those look like in Python? Powell's 2001 paper on the importance of modeling is no less relevant now than it was when published. I still many of his ideas in my spreadsheet based modeling course.

Quant Econ example

Teach relevant Python to business students using familiar and relevant modeling examples.
Give sense of software development process and design decisions
Model a learning journey
Teach students to think and act like a software developer, not just an analyst

Created the ``Excel with Python'' (EwP) module as part

While spreadsheets have certainly changed and expanded in their capabilities (PowerPivot, um...) since the 90's...

Irony of Mgt Sci crisis of 1990's where course was dropped by AACSB and ignored by other business disciplines to an era where analytics has captured the day and now AI threatens even that.

%TODO: Integrated or separate lit review section?

>>>>

Despite numerous calls for and predictions of the eradication of spreadsheets from the business world, Excel and other spreadsheets are alive and well. Yes, spreadsheets have well known limitations. Spreadsheets are not databases and probably should not be used as one, especially for relational data models. There are better data visualization and exploration tools. There are better dashboarding tools that are more easily maintained in a production environment. Spreadsheet errors can be a huge problem. Version control and documentation present challenges. There are data size limitations and external data links present their own set of challenges. But, Excel is widely used for building and using models of business problems
to explore the impact of various model inputs on key outputs. This ``what if?'' type modeling is facilitated by Excel's flexibility, numerous built in functions, integrated sensitivity analysis via \href{https://support.microsoft.com/en-us/office/calculate-multiple-results-by-using-a-data-table-e95e2487-6ca6-4413-ad12-77542a5ea50b}{Data Tables},  \href{https://support.microsoft.com/en-us/office/use-goal-seek-to-find-the-result-you-want-by-adjusting-an-input-value-320cb99e-f4a4-417f-b1c3-4f369d6e66c7}{Goal
	Seek} or break-even analysis and visualization capabilities.

Nevertheless, the analytics world has been changing. The advent of programmatic analytics using tools like R and Python have changed the practice of analytics as well as the teaching landscape \cite{bibid}. Limitations of spreadsheets with respect to reproducibility and automation have certainly played a role.  Additionally, both R and Python are free, open source, and have a rich set of analytics related libraries that have contributed to their growing popularity. Python, being a general purpose programming language, also offers advantages when moving models and analytical analysis pipelines from the prototype stage into production. Enormous data science ecosystems have emerged around both R and Python providing a wealth of learning resources for both new and experienced analytics students and professionals.

Numerous R and Python based data science courses have been developed and offered at universities and through online learning platforms such as Coursera, EdX, Udemy and others. Some of these courses have made a large amount of their course content freely available for self-learners and for other educators. At our institution, we have had a spreadsheet based modeling course in place since 2001. In 2015, we created a course called Practical Computing for Data Analytics (PCDA) that introduced both R and Python (along with a little Linux) for use in data analytics work. The course is taught using a combination of R Markdown documents \cite{bibid} and Jupyter notebooks \cite{bibid}. Both of these support \textit{literate computing} \cite{Jupyter_proposal} and allow us to create educational narratives that support iterative exploration and development. The ability to weave standard language with formal programming languages has made notebook computing a defacto standard in the data science education community. All of the course content, including many Jupyter notebooks and supporting screencasts, is freely available from a publicly accessible course web site \url{URL}.

The PCDA course, like many data science courses, focuses mostly on data wrangling, exploratory data analysis (EDA), and statistical/machine learning based predictive models. It does include a Python based simulation of the Monte Hall 3-door problem. In teaching EDA using tools like ggplot2, students are challenged to think about how they would go about trying to use Excel to create some of the plots that are easy to do with R and Python based plotting libraries. There is also some limited coverage of using Python to open Excel files and do some sheet manipulation. However, there was no real attempt to do the kind of modeling that was the focus of spreadsheet based modeling textbooks.

Python has continued to make inroads to analytical territory that was formerly dominated by Excel. The widely used Python data analysis library, pandas, was created by Wes McKinney while working as a quantitative financial analyst \cite{bibid}. Python is widely used for data wrangling, visualization and predictive modeling using various flavors of regression, tree based models and neural networks. The blog \href{https://pbpython.com/}{Practical Business Python} written primarily by Chris Moffitt contributed to making Python more accessible to business analytics practitioners through high quality posts that were extremely practical and useful. But, it got me to thinking. What about those things for which Excel is well suited such as building formula based models and doing sensitivity analysis on these models? What would those look like in Python? The importance of modeling, articulated so well in the pages of this journal \cite{powellTeachingModelingManagement2001}, has not abated. What role does Python have to play in supporting the kind of modeling taught for years in our business schools?

In 2021, I published a series of blog posts that explored this question and formed the basis of the teaching module described in this paper. During the development of these posts, I came across the \href{https://quantecon.org/}{Quant Econ: open source code for economic modeling} project which \quote{is a nonprofit organization dedicated to development and documentation of open source computational tools for economics, econometrics, and decision making}. This is a mature project, run by renowned economists \cite{bibid}, with significant support. They have developed and released numerous high quality Jupyter notebooks exploring a range of computational economic and financial modeling topics. Here was exactly the kind Python for business usage I had been seeking. With this as inspiration, I decided to turn my blog post series into a teaching module called Excel with Python (EwP).

The EwP module is not as polished nor as advanced as the Quant Econ materials, but does add to the growing body of freely available teaching materials aimed at Python for business modeling. My goals for EwP are also a little different and include:

\begin{itemize}
	\item Teach relevant Python to business students using familiar and relevant modeling examples,
	\item Give sense of software development process and design decisions
	\item Model a learning journey
	\item Teach students to think and act like a software developer, not just an analyst
\end{itemize}

One criticism that has been levied against data science programs has been the lack of attention to basic software engineering principles. The image of newly minted grads developing spaghetti coded Jupyter notebooks not nearly ready for being put into production has reached meme like status. I hoped to address some of these issues in the EwP module. Sneakily, students would be exposed to some simple software engineering concepts during the process of building and analyzing familiar models. This is really no different than the integration of advanced Excel functions and techniques with spreadsheet modeling tutorials. Just as spreadsheet modeling students were able to easily wow their coworkers and managers with INDEX-MATCH and intermediate Pivot Table skills, I hoped that my students ability to properly document a Python function and use git and GitHub would pay off in the workplace.
